---
title: "BioSCape Photo Processing"
format: html
editor: visual
---

```{r}
# Load necessary libraries
library(jsonlite)  # for handling JSON data
library(exifr)     # for extracting EXIF metadata from image files
library(lubridate) # for handling date and time data
library(stringr)   # for string operations
library(googlesheets4)
library(tidyverse)
library(doParallel)
library(googledrive)
library(readxl)
library(knitr)
library(piggyback)
library(leaflet)
library(magick)

tag=paste0("vegphoto_v",format(today(),"%Y%m%d"))
pb_new_release("BioSCape-io/BioSCape-terrestrial",tag=tag)
```

# Process combined photo spreadsheet

## Download current version of spreadsheet

```{r}
######### User settings
if (Sys.getenv("USER") == "jasper") {gmail = "jasper.slingsby@uct.ac.za"}
if (Sys.getenv("USER") == "adamw") {gmail = "adamw@buffalo.edu"}

# Define the URL or key of the Google Sheets
fname="data/Rarefaction_FA.xlsx"
sheet_url="https://docs.google.com/spreadsheets/d/13kJ7O4PbH6bM-Y89wGf_tbxs-R1THXlb/edit?usp=sharing&ouid=100268570982256677112&rtpof=true&sd=true"

drive_download(sheet_url, path = fname, overwrite = TRUE)
```

## Get sheet names

```{r}
sheets = excel_sheets(fname)
# Read sheets and do a little cleaning for irregularites

photo_all=foreach(i=sheets,.combine=bind_rows) %do% {
  read_xlsx(fname,i) %>% 
#    select(modify_date) |>
    mutate(
      fdate = gsub(":", "-", substring(modify_date, 1, 10)), # get just date part and replace : with -
      fdate2 = paste(fdate, substring(modify_date, 12)),       # Combine the formatted date part with the time part
      datetime = as_datetime(fdate2),       # Convert to datetime and extract the date using as_date
      date = date(datetime),       # extract just date
      time = hms(format(datetime, "%H:%M:%S")),  # get just time part from the datetime
      gps_latitude=as.numeric(gps_latitude),
      gps_longitude=as.numeric(gps_longitude),
      gps_altitude=as.numeric(gps_altitude),
      gps_position_error=as.numeric(gps_position_error),
      modify_date=as_datetime(modify_date),
      plot_number=as.character(plot_number),
      photo_type=case_when(
        grepl("plot|S|N|W|E|start|rarefaction|end|stop",fixed=F,genus) ~ "plot",
        T ~ "species"
        ),
      ) |>
    select(-fdate,-fdate2) |>  # drop temporary columns
    select(-exposure_time) |> #drop exposure time due to irregularity and not needed.
    filter(!is.na(gps_longitude)&!is.na(gps_latitude)) |> # this drops a few random screenshots
    st_as_sf(coords=c("gps_longitude","gps_latitude"),
           crs=4326) 
    } 

```

### Explore spatial data intersection

```{r}
# download spatial data
pb_download(file="BioScapeSpatialProducts2024_11_12.zip",dest = "data")
unzip("data/BioScapeSpatialProducts2024_11_12.zip",exdir = "data/spatial")
plots=read_sf("data/spatial/BioSCapeVegCenters2024_11_12.shp")
homogeneous=read_sf("data/spatial/BioSCapeVegPolys2024_11_12.shp")


dists=st_distance(photo_all, plots) %>% 
  tibble() |>
  rowwise() |>
  mutate(dist_to_plot_m=apply(.,1,min),
         min_which=apply(.,1,which.min),
         distplot=plots$BScpPID[min_which]) |>
  select(distplot,dist_to_plot_m)

photo_all2 <-  photo_all |>
     st_transform(st_crs(plots)) |>
     st_join(select(plots,BScpPID),st_nearest_feature) |>
     bind_cols(dists)


if(F){
select(photo_all2,BScpPID,distplot,dist_to_plot_m)|> View()
identical(photo_all2$distplot,photo_all2$BScpPID) #check if the nearest plot is the same as the plot in the spatial data)
}
# Plot histogram of distance to plot
    photo_all2 |>
    filter(dist_to_plot_m<50) |>
    ggplot(aes(x=dist_to_plot_m))+ 
    geom_histogram(binwidth=0.1)

```


# Plot identification issues
The following table shows plots that were labeled differently in the photo data compared to the spatial data. This could be due to a mislabeling in the photo data or a mislabeling in the spatial data. 

```{r}
select(photo_all2,plot_number,BScpPID)|>
    st_set_geometry(NULL)|>
    mutate(BScpPID2=as.numeric(gsub("T","",BScpPID)),
           plot_number=as.numeric(plot_number))|>
    filter(is.na(plot_number) | BScpPID2!=plot_number)|>
    distinct()|>
    kable()

```


## Upload photo dataset to github release

```{r}
photo_all_file=paste0("data/photo_all_",tag,".csv")
write_csv(photo_all2,photo_all_file)

# upload to github release is manual - run the pb_upload command if desired
if (F) pb_upload(photo_all_file,tag=tag)

```

## Rename and organize all photos

```{r}

photo_all3 <- photo_all2 |>
  mutate(
    newfile=paste0(BScpPID,"/",BScpPID,"_",photo_type,"_",format(datetime,"%Y%m%d_%H%M%S"),".jpeg"),
    thumbfile=paste0(BScpPID,"/",BScpPID,"_",photo_type,"_",format(datetime,"%Y%m%d_%H%M%S"),"_thumb.jpeg")
    )

output_folder="data/photo_out/"

# Define the standard size (e.g., 800x600 pixels)
standard_width <- 800
standard_height <- 600

foreach(i=1:nrow(photo_all3)) %do% {
 input_file= file.path("/Users/adamw/Library/CloudStorage/GoogleDrive-adammichaelwilson@gmail.com/Shared drives",photo_all3$folder[i],photo_all3$filename[i])
 output_file=file.path(outdir,photo_all3$newfile[i])
 output_thumb=file.path(outdir,photo_all3$thumbfile[i])
 
 if(!dir.exists(dirname(output_file))) dir.create(dirname(output_file),recursive = T)
 
 img <- image_read(input_file)
# Resize the image to the standard size
 thumb <- image_resize(img, paste0(standard_width, "x", standard_height))
  
    # Save the image as .jpeg
  image_write(img, output_file, format = "jpeg")
  image_write(thumb, output_thumb, format = "jpeg")
  
  # Print progress
  cat("Converted:", input_file, "->", output_file, "\n")

  }



#View(photo_all3)

```{r}
kable(head(photo_all))
```



# Map of plot locations with photos
```{r, echo=F}

# Create a leaflet map
leaflet() %>%
  #addProviderTiles("Esri.WorldImagery",tileOptions(maxZoom = 50)) %>%
  addWMSTiles(
    baseUrl = "http://aerial.openstreetmap.org.za/ngi-aerial/{z}/{x}/{y}.jpg",  # Replace with your WMS URL
    layers = "ngi-aerial",  # Replace with the desired WMS layer name
    options = WMSTileOptions(format = "image/jpg", transparent = TRUE,maxZoom=22),
    attribution = "© SA CDNGI Geospatial 25cm"
  ) %>% 
  addPolygons(data=homogeneous,fillColor="blue",fillOpacity=0.2,
              popup = ~paste("Region: ",homogeneous$Region,
                             "<br>Plot: ",homogeneous$BScpPID)) %>%
  addCircleMarkers(data=photo_all2,#lng=photo_all2$gps_longitude,lat=photo_all2$gps_latitude,
                   popup=~paste("Plot:",photo_all2$plot_number,"<br>Photo Description:",photo_all2$description,"<br>Genus:",photo_all2$genus,"<br>Species:",photo_all2$species,
                                "<br> <img src='data/photo_out/",photo_all3$thumbfile , "' style='width:150px;height:auto;'>"),
                   color="red",radius=1) %>%
  addCircleMarkers(data=plots,lng=st_coordinates(plots)[,"X"], lat=st_coordinates(plots)[,"Y"],
                   popup=~BScpPID,
                   color="blue",radius=3)

 
```


## Process Rarefaction photos

```{r}
photo_rarefaction <- photo_all |>
  group_by(plot_number, rarefaction_replicate) |>
  filter(rarefaction_photo==1) |>
  mutate(
    rarefaction_time=datetime-min(datetime),
    rarefaction_min=as.numeric(rarefaction_time)/60,
    count=row_number()-1,
    plot_number_replicate=paste(plot_number,rarefaction_replicate,sep=".")
    ) |>
  select(plot_number,plot_number_replicate,genus, species, datetime,time,rarefaction_time,rarefaction_min,rarefaction_replicate,count)
```

