---
title: "BioSCape Photo Processing"
format: gfm
editor: visual
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Load necessary libraries
library(jsonlite)  # for handling JSON data
library(exifr)     # for extracting EXIF metadata from image files
library(lubridate) # for handling date and time data
library(stringr)   # for string operations
library(googlesheets4)
library(tidyverse)
library(doParallel)
library(googledrive)
library(readxl)
library(knitr)
library(piggyback)
library(leaflet)
library(magick)
library(DT)
library(sf)

tagdate=paste0("v",format(today(),"%Y%m%d"))
tag=paste0("vegphoto_",tagdate)
#tag = "vegphoto_v20241126"
if(F){
pb_new_release("BioSCape-io/BioSCape-terrestrial",tag=tag)
}
```

# Process combined photo spreadsheet

## Download current version of spreadsheet

```{r, message=FALSE, warning=FALSE, echo=FALSE}
######### User settings
if (Sys.getenv("USER") == "jasper") {gmail = "jasper.slingsby@uct.ac.za"}
if (Sys.getenv("USER") == "adamw") {gmail = "adamw@buffalo.edu"}

# Define the URL or key of the Google Sheets
fname="data/Rarefaction_FA.xlsx"
sheet_url="https://docs.google.com/spreadsheets/d/13kJ7O4PbH6bM-Y89wGf_tbxs-R1THXlb/edit?usp=sharing&ouid=100268570982256677112&rtpof=true&sd=true"

drive_download(sheet_url, path = fname, overwrite = TRUE)
```

## Get sheet names

```{r, message=FALSE, warning=FALSE, echo=FALSE}
sheets = excel_sheets(fname)
# Read sheets and do a little cleaning for irregularites

sheets=sheets[-1] # Drop the old ross spreadsheet

photo_all=foreach(i=sheets,.combine=bind_rows) %do% {
  read_xlsx(fname,i) %>% 
#    select(modify_date) |>
    mutate(
      fdate = gsub(":", "-", substring(modify_date, 1, 10)), # get just date part and replace : with -
      fdate2 = paste(fdate, substring(modify_date, 12)),       # Combine the formatted date part with the time part
      datetime = as_datetime(fdate2),       # Convert to datetime and extract the date using as_date
      date = date(datetime),       # extract just date
      time = hms(format(datetime, "%H:%M:%S")),  # get just time part from the datetime
      gps_latitude=as.numeric(gps_latitude),
      gps_longitude=as.numeric(gps_longitude),
      gps_altitude=as.numeric(gps_altitude),
      gps_position_error=as.numeric(gps_position_error),
      modify_date=as_datetime(modify_date),
      plot_number=as.character(plot_number),
      photo_type=case_when(
        grepl("plot|S|N|W|E|start|rarefaction|end|stop",fixed=F,genus) ~ "plot",
        T ~ "species"
        ),
      ) |>
    select(-fdate,-fdate2) |>  # drop temporary columns
    select(-exposure_time) |> #drop exposure time due to irregularity and not needed.
    filter(!is.na(gps_longitude)&!is.na(gps_latitude)) |> # this drops a few random screenshots
    st_as_sf(coords=c("gps_longitude","gps_latitude"),remove=FALSE,
           crs=4326) 
    } 

```

### Explore spatial data intersection

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# download spatial data
pb_download(file="BioScapeSpatialProducts2024_11_12.zip",dest = "data")
unzip("data/BioScapeSpatialProducts2024_11_12.zip",exdir = "data/spatial")
```


```{r, message=FALSE, warning=FALSE, echo=FALSE}
plots=read_sf("data/spatial/BioSCapeVegCenters2024_11_12.shp")
homogeneous=read_sf("data/spatial/BioSCapeVegPolys2024_11_12.shp")


dists=st_distance(photo_all, plots) %>% 
  tibble() |>
  rowwise() |>
  mutate(dist_to_plot_m=apply(.,1,min),
         min_which=apply(.,1,which.min),
         distplot=plots$BScpPID[min_which]) |>
  select(distplot,dist_to_plot_m)|>
  mutate(dist_to_plot_m=round(dist_to_plot_m,0))

photo_all2 <-  photo_all |>
     st_transform(st_crs(plots)) |>
     st_join(select(plots,BScpPID),st_nearest_feature) |>
     bind_cols(dists)

if(F){

  select(photo_all2,BScpPID,distplot,dist_to_plot_m)|> View()
  identical(photo_all2$distplot,photo_all2$BScpPID) #confirm output from two methods above is the same

    photo_all2 |>
#    filter(dist_to_plot_m<50) |>
    ggplot(aes(x=dist_to_plot_m))+ 
    geom_histogram()#binwidth=0.1)

  }
# Plot histogram of distance to plot

```


# Plot identification issues
The following table shows plots that were labeled differently in the photo data compared to the spatial data. This could be due to a mislabeling in the photo data or a mislabeling in the spatial data. 

```{r, message=FALSE, warning=FALSE, echo=FALSE}
select(photo_all2,plot_number,BScpPID,dist_to_plot_m)|>
    st_set_geometry(NULL)|>
    mutate(BScpPID2=as.numeric(gsub("T","",BScpPID)),
           plot_number=as.numeric(plot_number))|>
    filter(is.na(plot_number) | BScpPID2!=plot_number)|>
    distinct()|>
    datatable()

```


## Upload photo dataset to github release

```{r, message=FALSE, warning=FALSE, echo=FALSE}
photo_all_file=paste0("data/vegphoto_all_",tagdate,".csv")
write_csv(photo_all2,photo_all_file)

# upload to github release is manual - run the pb_upload command if desired
if (F) pb_upload(photo_all_file,tag=tag)

```

## Rename and organize all photos

```{r, eval=F}

photo_all3 <- photo_all2 |>
  mutate(
    newfile=paste0(BScpPID,"/",BScpPID,"_",photo_type,"_",format(datetime,"%Y%m%d_%H%M%S"),".jpeg"),
    thumbfile=paste0(BScpPID,"/",BScpPID,"_",photo_type,"_",format(datetime,"%Y%m%d_%H%M%S"),"_thumb.jpeg")
    )

output_folder="data/photo_out/"

# Define the standard size (e.g., 800x600 pixels)
standard_width <- 800
standard_height <- 800

foreach(i=1:nrow(photo_all3)) %do% {
 input_file= file.path("/Users/adamw/Library/CloudStorage/GoogleDrive-adammichaelwilson@gmail.com/Shared drives",photo_all3$folder[i],photo_all3$filename[i])
 output_file=file.path(output_folder,photo_all3$newfile[i])
 output_thumb=file.path(output_folder,photo_all3$thumbfile[i])
 
 if(!dir.exists(dirname(output_file))) dir.create(dirname(output_file),recursive = T)
 
 img <- image_read(input_file)
# Resize the image to the standard size
 thumb <- image_resize(img, paste0(standard_width, "x", standard_height))
  
    # Save the image as .jpeg
  image_write(img, output_file, format = "jpeg")
  image_write(thumb, output_thumb, format = "jpeg")
  
  # update exif information in the new files

    cmd <- list(
    paste0("-overwrite_original -Description='original_path=", photo_all3$folder[i],"/",photo_all3$filename[i],
           " description=",photo_all3$description[i],
           " botanist=",photo_all3$botanist[i],"'"),    # original Description
    paste0("-Subject='", photo_all3$genus[i]," ",photo_all3$species[i],"'"),            # Genus Species
    paste0("-Title='BioSCape ", photo_all3$photo_type[i]," photo from plot ",photo_all3$BScpPID[i],"'"),             # photo type
    paste0("-Keywords='BioSCape,",photo_all3$botanist[i],"'")     # Original filename
  )
  
  # Update metadata in both output images
  exiftool_call(args = cmd,fnames=output_file)
  exiftool_call(args = cmd,fnames=output_thumb)
  
  #remove the automatic copies
}
```

# Upload Photos to Github Release

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# upload to github release is manual - run the pb_upload command if desired

if (F) pb_upload(output_folder,tag=tag)



```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
kable(head(photo_all))
```


