---
title: "BioSCape Photo Processing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show the code"
---

Required libraries were loaded and a versioned release tag was generated from the current date to label all output files and GitHub releases produced by this workflow.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Load necessary libraries
library(jsonlite)  # for handling JSON data
library(exifr)     # for extracting EXIF metadata from image files
library(lubridate) # for handling date and time data
library(stringr)   # for string operations
library(googlesheets4)
library(tidyverse)
library(doParallel)
library(googledrive)
library(readxl)
library(knitr)
library(piggyback)
library(leaflet)
library(magick)
library(DT)
library(sf)

tagdate=paste0("v",format(today(),"%Y%m%d"))
tag=paste0("vegphoto_",tagdate)
#tag = "vegphoto_v20241126"
if(F){
pb_new_release("BioSCape-io/BioSCape-terrestrial",tag=tag)
}
```

# Process combined photo spreadsheet

## Download current version of spreadsheet

The master photo metadata spreadsheet (`Rarefaction_FA.xlsx`) was downloaded from Google Drive. This file contains one worksheet per botanist, each populated in the field using a standardised template capturing filename, GPS coordinates, timestamp, species identification, and photo type flags.

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
######### User settings
if (Sys.getenv("USER") == "jasper") {gmail = "jasper.slingsby@uct.ac.za"}
if (Sys.getenv("USER") == "adamw") {gmail = "adamw@buffalo.edu"}

# Define the URL or key of the Google Sheets
fname="data/Rarefaction_FA.xlsx"
sheet_url="https://docs.google.com/spreadsheets/d/13kJ7O4PbH6bM-Y89wGf_tbxs-R1THXlb/edit?usp=sharing&ouid=100268570982256677112&rtpof=true&sd=true"

if(F) drive_download(sheet_url, path = fname, overwrite = TRUE)
```

## Parse and combine all botanist sheets

Each botanist worksheet was read and combined into a single dataset. Rows lacking a timestamp were discarded as incomplete entries. The `modify_date` field (formatted `YYYY:MM:DD HH:MM:SS` by the camera firmware) was re-formatted to ISO-8601 and parsed into separate `datetime`, `date`, and `time` columns. GPS coordinate fields were coerced to numeric; cells containing non-numeric text from Excel were silently converted to `NA`. Photos were classified as either `"plot"` (cardinal direction shots, plot overview, or rarefaction sequence markers) or `"species"` (individual taxon records) based on the content of the `genus` field. Cardinal directions were matched as exact single-character strings to avoid false positives within genus names. Records with missing GPS coordinates (e.g. screenshots) were excluded, and the resulting table was converted to an `sf` point layer in WGS 84 (EPSG:4326).

```{r, message=FALSE, warning=FALSE, echo=FALSE}
sheets = excel_sheets(fname)
# Read sheets and do a little cleaning for irregularites

sheets=sheets[-1] # Drop the old ross spreadsheet

photo_all=foreach(i=sheets,.combine=bind_rows) %do% {
  read_xlsx(fname,i) |>
    # Drop rows with no timestamp — these are incomplete entries and cannot be
    # placed on a timeline; also prevents unparseable strings downstream.
    filter(!is.na(modify_date)) |>
    mutate(
      # modify_date arrives as a character from Excel: "YYYY:MM:DD HH:MM:SS"
      # Replace the colons in the date portion with dashes before parsing.
      fdate    = gsub(":", "-", substring(modify_date, 1, 10)),
      fdate2   = paste(fdate, substring(modify_date, 12)),
      datetime = as_datetime(fdate2),
      date     = date(datetime),
      time     = hms(format(datetime, "%H:%M:%S")),
      # GPS fields can arrive as character "NA" from Excel; suppress coercion
      # warnings because NA → NA is the intended behaviour here.
      gps_latitude      = suppressWarnings(as.numeric(gps_latitude)),
      gps_longitude     = suppressWarnings(as.numeric(gps_longitude)),
      gps_altitude      = suppressWarnings(as.numeric(gps_altitude)),
      gps_position_error = suppressWarnings(as.numeric(gps_position_error)),
      modify_date  = as_datetime(modify_date),
      plot_number  = as.character(plot_number),
      # Classify plot-level vs species photos based on the genus field.
      # Cardinal directions (S, N, W, E) are matched as whole words only to avoid
      # false positives inside genus names (e.g. "Erica" contains "E").
      photo_type = case_when(
        grepl("^(plot|start|rarefaction|end|stop)$", genus, ignore.case = TRUE) ~ "plot",
        grepl("^[SNWE]$", genus)                                                 ~ "plot",
        TRUE                                                                     ~ "species"
      )
    ) |>
    select(-fdate, -fdate2) |>       # drop temporary date-parsing columns
    select(-exposure_time) |>        # drop — inconsistent type across sheets
    filter(!is.na(gps_longitude) & !is.na(gps_latitude)) |>  # drop screenshots with no GPS
    st_as_sf(coords = c("gps_longitude", "gps_latitude"), remove = FALSE,
             crs = 4326)
}

```

### Assign photos to vegetation plots

BioSCape vegetation plot centroids and homogeneous-patch polygons were loaded from the versioned spatial products file. The Euclidean distance from each photo to every plot centroid was computed in the projected coordinate system of the plot layer. Each photo was assigned to its nearest plot; assignments were nullified for photos located more than 1,000 m from the nearest centroid, as these are unlikely to represent intended plot observations.

```{r, message=FALSE, warning=FALSE, echo=FALSE}

# download spatial data 
# pb_download(file="BioScapeSpatialProducts2024_11_12.zip",dest = "data") 
# unzip("data/BioScapeSpatialProducts2024_11_12.zip",exdir = "data/spatial")

plots=read_sf("data/spatial/BioSCapeVegCenters2024_11_12.shp")
homogeneous=read_sf("data/spatial/BioSCapeVegPolys2024_11_12.shp")


dists=st_distance(photo_all, plots) %>% 
  tibble() |>
  rowwise() |>
  mutate(dist_to_plot_m=apply(.,1,min),
         min_which=apply(.,1,which.min),
         BScpPID=plots$BScpPID[min_which]) |>
  select(BScpPID,dist_to_plot_m)|>
  mutate(dist_to_plot_m=round(dist_to_plot_m,0))

photo_all2 <-  photo_all |>
     st_transform(st_crs(plots)) |>
#     st_join(select(plots,BScpPID),st_nearest_feature) |> #alternative method not used
     bind_cols(dists) |> 
  mutate(BScpPID=ifelse(dist_to_plot_m>1000,NA,BScpPID)) # drop linked plots if taken more than 5km from plot centroid.

if(F){

  photo_all2 |> 
  select(photo_all2,BScpPID,distplot,dist_to_plot_m)|> View()
  identical(photo_all2$distplot,photo_all2$BScpPID) #confirm output from two methods above is the same

    photo_all2 |>
#    filter(dist_to_plot_m<50) |>
    ggplot(aes(x=dist_to_plot_m))+ 
    geom_histogram()#binwidth=0.1)

  }
# Plot histogram of distance to plot

if(F){
  photo_all2 |>
    datatable()

}

```

# Plot identification issues

Photos were cross-referenced against the spatial plot database by comparing the plot number recorded in the botanist spreadsheet (`plot_number`) with the nearest-neighbour plot ID (`BScpPID`) derived from GPS coordinates. Discrepancies between these two fields are tabulated below and may reflect data-entry errors, plot renumbering, or photos taken at an adjacent plot. Each case should be reviewed before downstream use.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
select(photo_all2,botanist,description,location,plot_number,BScpPID,dist_to_plot_m)|>
    st_set_geometry(NULL)|>    
    mutate(BScpPID2=as.numeric(gsub("T","",BScpPID)),
           plot_number=as.numeric(gsub("T","",plot_number)))|>
    filter(is.na(plot_number) | BScpPID2!=plot_number)|>
    group_by(BScpPID2) |>
    summarize(n_photos=n(),
              botanists = paste(unique(botanist),collapse=","), 
              plots = paste(unique(plot_number),collapse=", "), 
              BScpPID2 = paste(unique(BScpPID2),collapse=", "), 
              mean_dist=mean(dist_to_plot_m)) |>
    distinct()|>
    datatable()

```

# Exploratory Summary

## Plot-level summary

Per-plot summary statistics were computed, including counts of photos by type (species, rarefaction, plot overview, iNaturalist-flagged), the number of unique taxa photographed, rarefaction effort, survey dates, and GPS quality indicators. The summary table was written to disk for use in downstream analyses.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
plot_summary <- photo_all2 |>
  st_drop_geometry() |>
  # Exclude photos with no matched plot before summarising
  filter(!is.na(BScpPID)) |>
  group_by(BScpPID) |>
  summarise(
    # Botanists who contributed photos to this plot
    botanist = paste(sort(unique(botanist)), collapse = ", "),

    # Total photo counts
    photo_total        = n(),
    photo_species      = sum(photo_type == "species", na.rm = TRUE),
    photo_rarefaction  = sum(rarefaction_photo == 1, na.rm = TRUE),
    photo_plot         = sum(photo_type == "plot", na.rm = TRUE),
    # Photos flagged for iNaturalist upload
    photo_inat         = sum(inat_photo == TRUE, na.rm = TRUE),
    # Rarefaction duplicates (same individual re-photographed in sequence)
    photo_rarefaction_duplicate = sum(
      rarefaction_duplicate == TRUE, na.rm = TRUE
    ),

    # Species richness: distinct genus+species combos from species-type photos,
    # excluding rows where genus is NA (unidentified / plot-level shots)
    n_species = n_distinct(paste(
      genus[photo_type == "species" & !is.na(genus)],
      species[photo_type == "species" & !is.na(genus)]
    )),

    # Maximum rarefaction replicate number — indicates depth of rarefaction
    # sequence completed for this plot
    n_rarefaction_replicates = sum(rarefaction_replicate, na.rm = TRUE),

    # Survey effort: number of unique dates photos were taken
    n_dates    = n_distinct(date),
    date_first = min(date, na.rm = TRUE),
    date_last  = max(date, na.rm = TRUE),

    # Spatial QC: median distance from photos to the official plot centroid (m)
    median_dist_to_plot_m = median(dist_to_plot_m, na.rm = TRUE),
    # GPS precision reported by the device
    mean_gps_error_m = round(mean(gps_position_error, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  # Replace -Inf / Inf that arise from max/min on all-NA groups
  mutate(
    across(where(is.numeric), \(x) ifelse(is.infinite(x), NA_real_, x))
  )

# Save to disk for use in downstream analyses
plot_summary_file <- paste0("data/plot_summary_", tagdate, ".csv")
write_csv(plot_summary, plot_summary_file)

plot_summary |>
  datatable(
    caption = "Plot-level photo summary",
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE
  )
```

## Global photo diagnostics

Dataset-level summary statistics were compiled to document overall data completeness, coverage, and quality. Key metrics include total photo count, number of matched plots, botanist contributions, temporal range, and mean per-plot photo density.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Per-plot photo counts (needed for mean ± SD)
per_plot_counts <- photo_all2 |>
  st_drop_geometry() |>
  filter(!is.na(BScpPID)) |>
  count(BScpPID, name = "n_photos")

per_plot_species_counts <- photo_all2 |>
  st_drop_geometry() |>
  filter(!is.na(BScpPID), photo_type == "species") |>
  count(BScpPID, name = "n_species_photos")

photo_diagnostics <- tibble(
  Metric = c(
    "Total photos (all sheets)",
    "Total matched plots",
    "Total unmatched photos (no plot ID)",
    "Distinct botanists",
    "Overall date range",
    "Photos missing coordinates (dropped pre-analysis)",
    "Photos unmatched to any plot (> 5 km from centroid)",
    "Rarefaction duplicate photos",
    "Mean photos per plot (± SD)",
    "Mean species photos per plot (± SD)",
    "Total unique species across all plots"
  ),
  Value = c(
    # Total rows in the dataset
    nrow(st_drop_geometry(photo_all2)),

    # Number of distinct plots with at least one matched photo
    n_distinct(
      photo_all2$BScpPID[!is.na(photo_all2$BScpPID)]
    ),

    # Photos that could not be matched to a plot
    sum(is.na(photo_all2$BScpPID)),

    # Distinct botanists
    n_distinct(photo_all2$botanist, na.rm = TRUE),

    # Date range as a string
    paste(
      min(photo_all2$date, na.rm = TRUE),
      "to",
      max(photo_all2$date, na.rm = TRUE)
    ),

    # Note: photos without GPS were filtered out during loading; count is
    # therefore zero post-filter but informative to document.
    "Filtered during data load (GPS required)",

    # Photos > 5 km from nearest plot centroid
    sum(is.na(photo_all2$BScpPID)),

    # Rarefaction duplicates flagged in the spreadsheet
    sum(photo_all2$rarefaction_duplicate == TRUE, na.rm = TRUE),

    # Mean ± SD photos per matched plot
    paste0(
      round(mean(per_plot_counts$n_photos), 1),
      " ± ",
      round(sd(per_plot_counts$n_photos), 1)
    ),

    # Mean ± SD species-type photos per matched plot
    paste0(
      round(mean(per_plot_species_counts$n_species_photos), 1),
      " ± ",
      round(sd(per_plot_species_counts$n_species_photos), 1)
    ),

    # Total unique species (genus + species) across all plots, species photos only
    n_distinct(paste(
      photo_all2$genus[photo_all2$photo_type == "species" &
                         !is.na(photo_all2$genus)],
      photo_all2$species[photo_all2$photo_type == "species" &
                           !is.na(photo_all2$genus)]
    ))
  )
)

kable(photo_diagnostics, col.names = c("Metric", "Value"),
      caption = "Global photo diagnostics")
```

## Upload photo dataset to GitHub release

The processed photo metadata table was written to a versioned CSV file. The sf geometry column was dropped prior to writing, as GPS coordinates are retained as explicit `gps_latitude` and `gps_longitude` columns. The file was optionally uploaded to the corresponding GitHub release for archival and downstream use.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
photo_all_file=paste0("data/vegphoto_all_",tagdate,".csv")
# Drop sf geometry before writing — the geometry column serialises as WKT and is
# not needed downstream (lat/lon columns are retained from the original data).
st_drop_geometry(photo_all2) |>
  write_csv(photo_all_file)

# upload to github release is manual - run the pb_upload command if desired
if (F) pb_upload(photo_all_file,tag=tag)

```

## Rename and organize all photos

Photos were copied from the shared Google Drive source folders into a standardised local directory structure (`data/photo_out/<BScpPID>/`). Each file was renamed according to the convention `<BScpPID>_<photo_type>_<YYYYMMDD_HHMMSS>.jpeg`. A thumbnail version (800 × 800 px) was generated alongside each full-resolution copy. EXIF metadata in both files was updated to embed the original file path, species identification, plot ID, and botanist name, providing a persistent record of provenance within the image files themselves. This chunk is set to `eval=F` and must be run manually after verifying the Google Drive paths.

```{r, eval=F}

photo_all3 <- photo_all2 |>
  mutate(
    newfile=paste0(BScpPID,"/",BScpPID,"_",photo_type,"_",format(datetime,"%Y%m%d_%H%M%S"),".jpeg"),
    thumbfile=paste0(BScpPID,"/",BScpPID,"_",photo_type,"_",format(datetime,"%Y%m%d_%H%M%S"),"_thumb.jpeg")
    )

output_folder="data/photo_out/"

# Define the standard size (e.g., 800x600 pixels)
standard_width <- 800
standard_height <- 800

foreach(i=1:nrow(photo_all3)) %do% {
 input_file= file.path("/Users/adamw/Library/CloudStorage/GoogleDrive-adammichaelwilson@gmail.com/Shared drives",photo_all3$folder[i],photo_all3$filename[i])
 output_file=file.path(output_folder,photo_all3$newfile[i])
 output_thumb=file.path(output_folder,photo_all3$thumbfile[i])
 
 if(!dir.exists(dirname(output_file))) dir.create(dirname(output_file),recursive = T)
 
 img <- image_read(input_file)
# Resize the image to the standard size
 thumb <- image_resize(img, paste0(standard_width, "x", standard_height))
  
    # Save the image as .jpeg
  image_write(img, output_file, format = "jpeg")
  image_write(thumb, output_thumb, format = "jpeg")
  
  # update exif information in the new files

    cmd <- list(
    paste0("-overwrite_original -Description='original_path=", photo_all3$folder[i],"/",photo_all3$filename[i],
           " description=",photo_all3$description[i],
           " botanist=",photo_all3$botanist[i],"'"),    # original Description
    paste0("-Subject='", photo_all3$genus[i]," ",photo_all3$species[i],"'"),            # Genus Species
    paste0("-Title='BioSCape ", photo_all3$photo_type[i]," photo from plot ",photo_all3$BScpPID[i],"'"),             # photo type
    paste0("-Keywords='BioSCape,",photo_all3$botanist[i],"'")     # Original filename
  )
  
  # Update metadata in both output images
  exiftool_call(args = cmd,fnames=output_file)
  exiftool_call(args = cmd,fnames=output_thumb)
  
  #remove the automatic copies
}
```

# Upload Photos to GitHub Release

The renamed photo folder was optionally uploaded to the versioned GitHub release. This step is run manually to avoid unintentional large uploads.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# upload to github release is manual - run the pb_upload command if desired

if (F) pb_upload(output_folder,tag=tag)

```

## Preview raw data

The first rows of the combined photo table are displayed for a quick sanity check of column names, data types, and representative values.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
kable(head(photo_all))
```