---
title: "BioSCape iNaturalist Upload Preparation"
author: "Adam M. Wilson, Jasper Slingsby, and the BioSCape Team"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show the code"
execute:
  echo: true
  warning: false
  message: false
---

## Summary

This script reads the cleaned photo dataset produced by `botanist_photo_processing.qmd` and prepares batch CSV files for upload to iNaturalist using the `workflow/dist/iNat_bulk_uploader.exe` tool.

Each row in the output represents one species observation at one plot (one row per species × plot combination). Photos of the same species at the same plot are collapsed into up to 10 `media_name_*` columns. Observations are split into batches of 50 rows for upload.

**Prerequisites:**

1.  Run `botanist_photo_processing.qmd` to produce the `photo_all_file` CSV.
2.  Run the photo renaming chunk in `botanist_photo_processing.qmd` (currently `eval=F`) so that the `newfile` column is populated with renamed photo paths.

```{r setup}
library(tidyverse)
library(piggyback)
library(DT)
library(knitr)

# -------------------------------------------------------
# User parameters — update these before running
# -------------------------------------------------------

# Version tag of the photo_all file from botanist_photo_processing.qmd
photo_version <- "v20250228"

# Number of observations per iNat batch CSV
batch_size <- 50

# Output folder for batch CSVs
batch_dir <- "data/inat_batches"

# iNat tag applied to every observation
inat_tag <- "BioSCape"

# Time zone string for iNat (see https://stackoverflow.com/a/13867044)
inat_timezone <- "Africa/Johannesburg"

# -------------------------------------------------------
# Derived paths
# -------------------------------------------------------
photo_all_file <- file.path("data", paste0("vegphoto_all_", photo_version, ".csv"))
```

## Load photo data

```{r load-data}
# Download from GitHub release if not present locally
if (!file.exists(photo_all_file)) {
  pb_download(
    file    = basename(photo_all_file),
    dest    = "data",
    tag     = paste0("vegphoto_", photo_version)
  )
}

photo_all <- read_csv(photo_all_file, show_col_types = FALSE)

# Validate that the newfile column exists and is populated.
# The 'newfile' column is created by the photo-renaming chunk in
# botanist_photo_processing.qmd (currently eval=F — must be run manually first).
if (!"newfile" %in% names(photo_all)) {
  stop(
    "Column 'newfile' not found in photo_all.\n",
    "Run the photo-renaming chunk in botanist_photo_processing.qmd first."
  )
}

n_missing_newfile <- sum(is.na(photo_all$newfile))
if (n_missing_newfile > 0) {
  warning(
    n_missing_newfile,
    " rows have NA in 'newfile'. These observations will have no media attached."
  )
}

glimpse(photo_all)
```

## Filter to species photos

```{r filter-species}
# Keep only photos tagged as species observations (not plot/cardinal/rarefaction shots).
# Also require non-NA genus, species, and plot ID so that the iNat taxon_name and
# location fields are always populated.
species_photos <- photo_all |>
  filter(
    photo_type == "species",
    !is.na(genus),
    !is.na(species),
    !is.na(BScpPID)
  ) |>
  mutate(
    taxon_name = paste(genus, species),
    # iNat requires date in DD/MM/YYYY format
    date_obs   = format(as.Date(date), "%d/%m/%Y"),
    # basename of the renamed photo file, used in media_name_* columns
    photo_basename = basename(newfile)
  )

cat("Species photos available for iNat upload:", nrow(species_photos), "\n")
cat("Unique taxon names:                      ", n_distinct(species_photos$taxon_name), "\n")
cat("Unique plots:                            ", n_distinct(species_photos$BScpPID), "\n")
```

## Collapse to one row per species × plot

```{r collapse-observations}
# The iNat bulk uploader expects one observation per row.
# We define one observation = one species at one plot.
# All photos of that species at that plot are spread across media_name_1 … media_name_10.
# GPS coordinates are averaged across photos; positional accuracy is the maximum.
# The description field concatenates all unique photo captions for context.

# Maximum number of media columns in the template
max_media <- 10

inat_obs <- species_photos |>
  group_by(BScpPID, taxon_name) |>
  summarise(
    date_obs    = first(date_obs),        # date of first photo of this species at plot
    latitude    = mean(gps_latitude,  na.rm = TRUE),
    longitude   = mean(gps_longitude, na.rm = TRUE),
    pos_acc     = max(gps_position_error, na.rm = TRUE),
    botanist    = paste(unique(na.omit(botanist)), collapse = ", "),
    description = paste(
      paste0("BioSCape Vegetation Plot=", first(BScpPID),
             "; Botanist=", paste(unique(na.omit(botanist)), collapse = "/"),
             "; Descriptions=", paste(unique(na.omit(description)), collapse = " | "))
    ),
    n_photos    = n(),
    # Collect photo basenames as a list, then keep up to max_media
    photos      = list(na.omit(photo_basename)[seq_len(min(n(), max_media))]),
    .groups = "drop"
  ) |>
  # Replace Inf/-Inf from max() on all-NA pos_acc with NA
  mutate(
    pos_acc = if_else(is.infinite(pos_acc), NA_real_, round(pos_acc, 1))
  )

cat("Total iNat observations (species × plot):", nrow(inat_obs), "\n")
cat("Batches of", batch_size, "needed:         ",
    ceiling(nrow(inat_obs) / batch_size), "\n")
```

## Build iNat template table

```{r build-template}
# Spread the photos list-column into media_name_1 … media_name_10 columns,
# then assemble the full iNat template with all required fields.

# Helper: extract nth element from a list column, returning NA if out of bounds
nth_photo <- function(photos_list, n) {
  map_chr(photos_list, ~ if (length(.x) >= n) .x[[n]] else NA_character_)
}

inat_table <- inat_obs |>
  mutate(
    # Sequential ID across the full upload table
    id           = row_number(),
    time_zone    = inat_timezone,
    tag_list     = inat_tag,
    location     = BScpPID,
    geoprivacy   = "open",
    # iNat observation field columns (empty for now — add BioSCape project field IDs here)
    field_id1    = NA_character_,
    field_value1 = NA_character_,
    field_id2    = NA_character_,
    field_value2 = NA_character_,
    # Spread photo basenames into separate columns
    media_name_1  = nth_photo(photos,  1),
    media_name_2  = nth_photo(photos,  2),
    media_name_3  = nth_photo(photos,  3),
    media_name_4  = nth_photo(photos,  4),
    media_name_5  = nth_photo(photos,  5),
    media_name_6  = nth_photo(photos,  6),
    media_name_7  = nth_photo(photos,  7),
    media_name_8  = nth_photo(photos,  8),
    media_name_9  = nth_photo(photos,  9),
    media_name_10 = nth_photo(photos, 10),
    # Batch assignment: ceiling(id / batch_size) gives the batch number
    batch         = ceiling(id / batch_size)
  ) |>
  # Select and order columns to match the iNat template exactly
  select(
    id, taxon_name, date_obs, time_zone, description, tag_list,
    latitude, longitude, location, pos_acc, geoprivacy,
    field_id1, field_value1, field_id2, field_value2,
    media_name_1, media_name_2, media_name_3, media_name_4, media_name_5,
    media_name_6, media_name_7, media_name_8, media_name_9, media_name_10,
    # keep batch and n_photos for QC — these are dropped when writing output CSVs
    batch, n_photos
  )
```

## QA check

```{r qa-check}
# Flag observations with potential data quality issues before writing output.

qa_flags <- inat_table |>
  mutate(
    flag_no_media    = is.na(media_name_1),   # no photo attached
    flag_no_taxon    = is.na(taxon_name) | taxon_name == "NA NA",
    flag_no_coords   = is.na(latitude) | is.na(longitude),
    flag_high_pos    = !is.na(pos_acc) & pos_acc > 100  # >100 m accuracy is suspicious
  ) |>
  filter(flag_no_media | flag_no_taxon | flag_no_coords | flag_high_pos) |>
  select(id, BScpPID, taxon_name, n_photos,
         flag_no_media, flag_no_taxon, flag_no_coords, flag_high_pos)

cat("Observations with QA flags:", nrow(qa_flags), "\n")

if (nrow(qa_flags) > 0) {
  qa_flags |>
    datatable(caption = "Observations with QA flags — review before uploading")
}
```

## Preview

```{r preview}
inat_table |>
  select(id, taxon_name, date_obs, BScpPID, latitude, longitude,
         n_photos, batch, media_name_1) |>
  slice(1:20) |>
  datatable(caption = "First 20 iNat observations (preview)")
```

## Batch summary

```{r batch-summary}
inat_table |>
  count(batch, name = "n_observations") |>
  kable(caption = paste("Batch summary —", batch_size, "observations per batch"))
```

## Write batch CSV files

```{r write-batches}
# Create output directory
if (!dir.exists(batch_dir)) dir.create(batch_dir, recursive = TRUE)

# iNat template columns only — exclude 'batch' and 'n_photos'
template_cols <- c(
  "id", "taxon_name", "date_obs", "time_zone", "description", "tag_list",
  "latitude", "longitude", "location", "pos_acc", "geoprivacy",
  "field_id1", "field_value1", "field_id2", "field_value2",
  "media_name_1", "media_name_2", "media_name_3", "media_name_4", "media_name_5",
  "media_name_6", "media_name_7", "media_name_8", "media_name_9", "media_name_10"
)

batch_ids <- sort(unique(inat_table$batch))

walk(batch_ids, function(b) {
  batch_file <- file.path(batch_dir,
                          paste0("batch_", sprintf("%02d", b), ".csv"))
  inat_table |>
    filter(batch == b) |>
    select(all_of(template_cols)) |>
    write_csv(batch_file, na = "")   # iNat uploader expects empty strings not "NA"
  cat("Wrote:", batch_file, "\n")
})

cat("\nAll batches written to:", batch_dir, "\n")
cat("Total batches:", length(batch_ids), "\n")
cat("Total observations:", nrow(inat_table), "\n")
```

## Next steps

1.  Copy the batch CSV files from `data/inat_batches/` to a folder alongside the renamed photos from `data/photo_out/`.
2.  Open `workflow/dist/iNat_bulk_uploader.exe`.
3.  Enter your iNaturalist username and password.
4.  Select a batch CSV and the photo folder, then submit.
5.  Check `workflow/dist/log.csv` after each batch for errors.
6.  Test with a small batch (`batch_01.csv`) first before uploading all batches.

After uploading, download the iNat observation export and join observation IDs back to `bioscape_veg_plot_species_v20241104.csv` using `taxon_name` + `BScpPID` (`location`).