---
title: "BioSCape iNaturalist Upload Preparation"
author: "Adam M. Wilson, Jasper Slingsby, and the BioSCape Team"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show the code"
execute:
  echo: true
  warning: false
  message: false
---

## Summary

This script reads the cleaned photo dataset produced by `botanist_photo_processing.qmd` and prepares batch CSV files for the iNaturalist bulk uploader (`workflow/dist/iNat_bulk_uploader.exe`; see `workflow/dist/readme.txt`).

Each output row is one corrected species at one plot. Photos of that species at that plot, including different raw spellings and rarefaction species photos, are collapsed into up to 10 `media_name_*` columns. Plot photos (cardinal shots, overviews, rarefaction markers, datasheets) are not uploaded. Observations are split into batches of 50.

The batch CSVs are written only after two things are true: `newfile` is populated by the photo-copy chunk, and both numeric observation-field ids below are set. Until then the notebook still builds the observation table so it can be reviewed.

After upload, the last section reads `data/inat_export.csv` and writes `iNatID` onto copies of the plot and quadrat species tables. The v20241104 files are left unchanged. If the export is not there yet, that section reports so and stops.

**Prerequisites:**

1. Run `botanist_photo_processing.qmd` to produce `data/vegphoto_all_<version>.csv`.
2. On iNaturalist, create two observation fields and paste their numeric ids (the number in `https://www.inaturalist.org/observation_fields/<id>`) into the parameters below. Name them exactly `BioSCape SiteCode_Plot` and `BioSCape Genus_Species_combo`. The uploader stores the number, not the name.
3. Run the photo-copy chunk in `botanist_photo_processing.qmd` (`eval=F`) so `newfile` is written back onto the vegphoto CSV.
4. The uploader is a Windows program. Copy one batch CSV and every JPEG it names into a single folder, run the exe there, and copy `workflow/dist/log.csv` aside after each batch. The log is replaced on the next run.

```{r setup}
library(tidyverse)
library(piggyback)
library(DT)
library(knitr)

# -------------------------------------------------------
# User parameters — update these before running
# -------------------------------------------------------

# Version tag of the photo_all file from botanist_photo_processing.qmd
photo_version <- "v20260923"

# Number of observations per iNat batch CSV
batch_size <- 50

# Output folder for batch CSVs
batch_dir <- "data/inat_batches"

# iNat tag applied to every observation
inat_tag <- "BioSCape"

# Time zone string for iNat (pytz name; see workflow/dist/readme.txt)
inat_timezone <- "Africa/Johannesburg"

# Numeric observation-field ids. Leave NA until the fields exist on iNaturalist.
# field_id in the batch CSV must be this number, not the field's display name.
inat_field_id_siteplot <- NA_integer_   # "BioSCape SiteCode_Plot"
inat_field_id_taxon    <- NA_integer_   # "BioSCape Genus_Species_combo"

# Display names used to find the same fields in an iNat export.
inat_field_name_siteplot <- "BioSCape SiteCode_Plot"
inat_field_name_taxon    <- "BioSCape Genus_Species_combo"

# -------------------------------------------------------
# Derived paths
# -------------------------------------------------------
photo_all_file <- file.path("data", paste0("vegphoto_all_", photo_version, ".csv"))
plot_species_file <- "data/bioscape_veg_plot_species_v20241104.csv"
quad_species_file <- "data/bioscape_veg_quadrat_species_v20241104.csv"
plot_inat_file <- "data/bioscape_veg_plot_species_v20241104_inat.csv"
quad_inat_file <- "data/bioscape_veg_quadrat_species_v20241104_inat.csv"
inat_export_file <- "data/inat_export.csv"

# Up to 10 media columns, matching the columns documented in the uploader readme.
max_media <- 10

field_id_chr <- function(x) {
  if (length(x) != 1 || is.na(x)) NA_character_ else as.character(x)
}
```

## Load photo data

```{r load-data}
# Download from GitHub release if not present locally
if (!file.exists(photo_all_file)) {
  pb_download(
    file    = basename(photo_all_file),
    dest    = "data",
    tag     = paste0("vegphoto_", photo_version)
  )
}

photo_all <- read_csv(photo_all_file, show_col_types = FALSE)

# The photo-copy chunk writes `newfile`. Until it has been run, keep an empty
# column so the observation table can still be reviewed. Batch CSVs are not
# written in that state.
if (!"newfile" %in% names(photo_all)) {
  message(
    "Column 'newfile' is not in the vegphoto file. ",
    "Run the photo-copy chunk in botanist_photo_processing.qmd before uploading. ",
    "The observation table will be built without media paths."
  )
  photo_all <- photo_all |> mutate(newfile = NA_character_)
}

n_missing_newfile <- sum(is.na(photo_all$newfile))
if (n_missing_newfile > 0) {
  message(n_missing_newfile, " photo rows have no newfile path.")
}

# `plot` in the vegphoto file is the same code as species `Plot` (for example T001).
if (!"plot" %in% names(photo_all)) {
  stop("Column 'plot' is missing from ", photo_all_file, ".")
}

veg_plot_species <- read_csv(plot_species_file, show_col_types = FALSE)
veg_plots <- veg_plot_species |>
  distinct(SiteCode, Plot, SiteCode_Plot)

photo_all <- photo_all |>
  rename(Plot = plot) |>
  left_join(veg_plots, by = "Plot")

glimpse(photo_all)
```

## Filter to species photos

```{r filter-species}
# Plot photos stay out. Rarefaction species photos stay in: many taxa have
# no other photo. Blank names, failed matches, and manual exclusions stay out.
# Genus-only rows upload as a genus. Corrected binomials upload even when that
# species is not on the plot's survey list.
species_photos <- photo_all |>
  filter(
    photo_type == "species",
    match_type %in% c("exact", "fuzzy", "spd_fill", "genus_only"),
    !is.na(corrected_genus),
    !is.na(Plot),
    match_type == "genus_only" | !is.na(corrected_species)
  ) |>
  mutate(
    Genus_Species_combo = if_else(
      match_type == "genus_only",
      NA_character_,
      str_squish(paste(corrected_genus, corrected_species))
    ),
    taxon_name = if_else(
      match_type == "genus_only",
      corrected_genus,
      Genus_Species_combo
    ),
    date_obs = format(as.Date(date), "%d/%m/%Y"),
    photo_basename = basename(newfile)
  )

n_no_site <- sum(is.na(species_photos$SiteCode_Plot))
if (n_no_site > 0) {
  warning(
    n_no_site,
    " species photos have a plot code with no SiteCode_Plot in the species table."
  )
}

cat("Species photos available for iNat upload:", nrow(species_photos), "\n")
cat("  binomials:  ", sum(!is.na(species_photos$Genus_Species_combo)), "\n")
cat("  genus-only: ", sum(is.na(species_photos$Genus_Species_combo)), "\n")
cat("Unique taxa:  ", n_distinct(species_photos$taxon_name), "\n")
cat("Unique plots: ", n_distinct(species_photos$Plot), "\n")
```

## Collapse to one row per species × plot

```{r collapse-observations}
# One observation = one corrected taxon at one plot. Raw spellings that
# corrected to the same name collapse together. Grouping is on `Plot`, not
# `SiteCode_Plot`, so photos whose site lookup failed are not merged across plots.
# Grouping is not on `Genus_Species_combo`: that column is blank for every
# genus-only photo, and different genera at one plot must stay separate.

nth_photo <- function(photos_list, n) {
  map_chr(photos_list, ~ if (length(.x) >= n) .x[[n]] else NA_character_)
}

inat_obs <- species_photos |>
  arrange(Plot, taxon_name, datetime) |>
  group_by(Plot, taxon_name) |>
  summarise(
    SiteCode            = first(SiteCode),
    SiteCode_Plot       = first(SiteCode_Plot),
    Genus_Species_combo = first(Genus_Species_combo),
    date_obs            = first(date_obs),
    latitude            = mean(gps_latitude, na.rm = TRUE),
    longitude           = mean(gps_longitude, na.rm = TRUE),
    pos_acc             = max(gps_position_error, na.rm = TRUE),
    botanist            = paste(unique(na.omit(botanist)), collapse = ", "),
    notes               = paste(unique(na.omit(description)), collapse = " | "),
    n_photos            = n(),
    photos              = list(head(na.omit(photo_basename), max_media)),
    .groups = "drop"
  ) |>
  mutate(
    pos_acc = if_else(is.infinite(pos_acc), NA_real_, round(pos_acc, 1)),
    notes   = if_else(notes == "", NA_character_, notes)
  )

n_dup <- inat_obs |>
  count(Plot, taxon_name) |>
  filter(n > 1) |>
  nrow()
if (n_dup > 0) {
  stop(n_dup, " plot × taxon combinations are still duplicated after collapse.")
}

n_over_cap <- sum(inat_obs$n_photos > max_media)
if (n_over_cap > 0) {
  message(
    n_over_cap,
    " observations have more than ", max_media,
    " photos. Extra photos are not included in the batch CSV."
  )
}

cat("iNat observations (one per plot × corrected taxon):", nrow(inat_obs), "\n")
cat("Batches of", batch_size, "needed:",
    ceiling(nrow(inat_obs) / batch_size), "\n")
```

## Build iNat template table

```{r build-template}
inat_table <- inat_obs |>
  mutate(
    id           = row_number(),
    time_zone    = inat_timezone,
    tag_list     = inat_tag,
    location     = SiteCode_Plot,
    geoprivacy   = "open",
    # Numeric field ids. Values are the strings written at upload so the
    # later join does not depend on iNat's current taxon name.
    field_id1    = field_id_chr(inat_field_id_siteplot),
    field_value1 = SiteCode_Plot,
    field_id2    = field_id_chr(inat_field_id_taxon),
    field_value2 = taxon_name,
    description  = paste0(
      "BioSCape Vegetation Plot=", SiteCode_Plot,
      "; SiteCode=", SiteCode,
      "; Plot=", Plot,
      "; Taxon=", taxon_name,
      "; Botanist=", botanist,
      "; Notes=", notes
    ),
    media_name_1  = nth_photo(photos,  1),
    media_name_2  = nth_photo(photos,  2),
    media_name_3  = nth_photo(photos,  3),
    media_name_4  = nth_photo(photos,  4),
    media_name_5  = nth_photo(photos,  5),
    media_name_6  = nth_photo(photos,  6),
    media_name_7  = nth_photo(photos,  7),
    media_name_8  = nth_photo(photos,  8),
    media_name_9  = nth_photo(photos,  9),
    media_name_10 = nth_photo(photos, 10),
    batch         = ceiling(id / batch_size)
  ) |>
  select(
    id, taxon_name, date_obs, time_zone, description, tag_list,
    latitude, longitude, location, pos_acc, geoprivacy,
    field_id1, field_value1, field_id2, field_value2,
    media_name_1, media_name_2, media_name_3, media_name_4, media_name_5,
    media_name_6, media_name_7, media_name_8, media_name_9, media_name_10,
    batch, n_photos, Plot, SiteCode, SiteCode_Plot, Genus_Species_combo
  )

obs_keys <- inat_table |>
  filter(!is.na(Genus_Species_combo), !is.na(SiteCode_Plot)) |>
  mutate(combo_key = str_to_lower(Genus_Species_combo)) |>
  distinct(SiteCode_Plot, combo_key)

no_obs <- veg_plot_species |>
  mutate(combo_key = str_to_lower(str_squish(Genus_Species_combo))) |>
  anti_join(obs_keys, by = c("SiteCode_Plot", "combo_key"))

cat("Plot-species rows with no iNat observation:", nrow(no_obs), "\n")
cat("Observations with no SiteCode_Plot:", sum(is.na(inat_table$SiteCode_Plot)), "\n")
```

## QA check

```{r qa-check}
# Until the photo-copy chunk has run, every observation lacks media. That is
# reported once above. Flag individual rows only after some newfile values exist.
newfile_ready_qa <- any(!is.na(species_photos$newfile))

qa_flags <- inat_table |>
  mutate(
    flag_no_media  = newfile_ready_qa & is.na(media_name_1),
    flag_no_taxon  = is.na(taxon_name) | taxon_name == "",
    flag_no_coords = is.na(latitude) | is.na(longitude),
    flag_no_site   = is.na(SiteCode_Plot),
    flag_high_pos  = !is.na(pos_acc) & pos_acc > 100
  ) |>
  filter(flag_no_media | flag_no_taxon | flag_no_coords | flag_no_site | flag_high_pos) |>
  select(id, location, taxon_name, n_photos,
         flag_no_media, flag_no_taxon, flag_no_coords, flag_no_site, flag_high_pos)

cat("Observations with QA flags:", nrow(qa_flags), "\n")

if (nrow(qa_flags) > 0) {
  qa_flags |>
    datatable(caption = "Observations with QA flags — review before uploading")
}
```

## Preview

```{r preview}
inat_table |>
  select(id, taxon_name, date_obs, location, latitude, longitude,
         n_photos, batch, media_name_1) |>
  slice(1:20) |>
  datatable(caption = "First 20 iNat observations (preview)")
```

## Batch summary

```{r batch-summary}
inat_table |>
  count(batch, name = "n_observations") |>
  kable(caption = paste("Batch summary —", batch_size, "observations per batch"))
```

## Write batch CSV files

```{r write-batches}
template_cols <- c(
  "id", "taxon_name", "date_obs", "time_zone", "description", "tag_list",
  "latitude", "longitude", "location", "pos_acc", "geoprivacy",
  "field_id1", "field_value1", "field_id2", "field_value2",
  "media_name_1", "media_name_2", "media_name_3", "media_name_4", "media_name_5",
  "media_name_6", "media_name_7", "media_name_8", "media_name_9", "media_name_10"
)

newfile_ready <- any(!is.na(species_photos$newfile))
fields_ready <- !is.na(inat_field_id_siteplot) && !is.na(inat_field_id_taxon)

if (!newfile_ready || !fields_ready) {
  missing <- c(
    if (!newfile_ready) "newfile (run the photo-copy chunk in botanist_photo_processing.qmd)",
    if (is.na(inat_field_id_siteplot)) "inat_field_id_siteplot",
    if (is.na(inat_field_id_taxon)) "inat_field_id_taxon"
  )
  message(
    "Batch CSVs were not written. Still needed: ",
    paste(missing, collapse = "; "),
    "."
  )
} else {
  if (!dir.exists(batch_dir)) dir.create(batch_dir, recursive = TRUE)

  # A blank media_name_1 makes the uploader abort the row. "none" is the
  # documented token for an observation that has no photo.
  batch_table <- inat_table |>
    mutate(media_name_1 = if_else(is.na(media_name_1), "none", media_name_1))

  batch_ids <- sort(unique(batch_table$batch))
  walk(batch_ids, function(b) {
    batch_file <- file.path(batch_dir, paste0("batch_", sprintf("%02d", b), ".csv"))
    batch_table |>
      filter(batch == b) |>
      select(all_of(template_cols)) |>
      write_csv(batch_file, na = "")
    cat("Wrote:", batch_file, "\n")
  })

  cat("\nAll batches written to:", batch_dir, "\n")
  cat("Total batches:", length(batch_ids), "\n")
  cat("Total observations:", nrow(batch_table), "\n")
}
```

## Next steps

The bulk uploader is the Windows program `workflow/dist/iNat_bulk_uploader.exe`.

1. Create the two observation fields on iNaturalist and set `inat_field_id_siteplot` and `inat_field_id_taxon` to the numeric ids from their field URLs.
2. Run the photo-copy chunk in `botanist_photo_processing.qmd`, then re-run this notebook so `newfile` is present and the batch CSVs are written.
3. Copy the JPEGs named in a batch into one folder. `media_name_*` is a basename, and the program looks only in the folder you select.
4. Run the exe, choose `batch_01.csv` and that folder, and submit. Copy `log.csv` out of the installation folder before the next batch; the program replaces it on every run.
5. After every batch has uploaded, export the observations to `data/inat_export.csv` and re-run the section below.

The program cannot add observations to an iNat project. `ABORT` in the `id` column stops a run. `config.ini` beside the exe still holds another person's paths; choose the CSV and photo folder in the program rather than editing that file.

## Write iNatID back to the species tables

`iNatID` is the observation id in `https://www.inaturalist.org/observations/<id>`, which is also the `id` parameter on an observation search. This section runs only when `data/inat_export.csv` is present. It writes new files and does not modify the v20241104 species tables.

A binomial observation is attached to every quadrat row of that species at that plot. Genus-only observations and binomials that are not in the survey have no matching row, so they do not fill `iNatID`. If more than one observation matches the same species row, the smallest id is kept and the extras are reported.

```{r backlink}
if (!file.exists(inat_export_file)) {
  message(
    "No iNat export at ", inat_export_file, ". ",
    "Upload the batches, export the observations, and re-run this section."
  )
} else {
  inat_export <- read_csv(inat_export_file, show_col_types = FALSE)

  # iNat prefixes observation-field columns as "field:<name>".
  field_column <- function(nm, field_name) {
    candidates <- c(field_name, paste0("field:", field_name))
    hit <- nm[str_to_lower(nm) %in% str_to_lower(candidates)]
    if (length(hit) != 1) {
      stop(
        "Expected one column for observation field '", field_name, "'. ",
        "Found: ", if (length(hit) == 0) "none" else paste(hit, collapse = ", "),
        ". Export columns: ", paste(nm, collapse = ", ")
      )
    }
    hit
  }

  if (!"id" %in% names(inat_export)) {
    stop("inat export has no 'id' column. Columns: ",
         paste(names(inat_export), collapse = ", "))
  }

  site_col  <- field_column(names(inat_export), inat_field_name_siteplot)
  taxon_col <- field_column(names(inat_export), inat_field_name_taxon)

  links <- inat_export |>
    transmute(
      iNatID = as.integer(id),
      SiteCode_Plot = str_squish(.data[[site_col]]),
      taxon_uploaded = str_squish(.data[[taxon_col]]),
      combo_key = str_to_lower(taxon_uploaded)
    ) |>
    filter(!is.na(iNatID), SiteCode_Plot != "", taxon_uploaded != "")

  # Only a binomial that exists on a survey row can receive an iNatID.
  # A genus-only value does not equal any Genus_Species_combo.
  survey_keys <- veg_plot_species |>
    distinct(SiteCode_Plot, Genus_Species_combo) |>
    mutate(combo_key = str_to_lower(str_squish(Genus_Species_combo))) |>
    select(SiteCode_Plot, combo_key)

  matched <- links |>
    inner_join(survey_keys, by = c("SiteCode_Plot", "combo_key"))

  extra <- matched |>
    group_by(SiteCode_Plot, combo_key) |>
    filter(n() > 1) |>
    arrange(iNatID) |>
    mutate(kept = iNatID == min(iNatID)) |>
    ungroup()
  if (nrow(extra) > 0) {
    warning(
      n_distinct(paste(extra$SiteCode_Plot, extra$combo_key)),
      " plot × species keys matched more than one iNat observation. ",
      "The smallest iNatID was kept."
    )
    extra |>
      filter(!kept) |>
      select(SiteCode_Plot, taxon_uploaded, iNatID) |>
      print(n = Inf)
  }

  links_one <- matched |>
    group_by(SiteCode_Plot, combo_key) |>
    summarise(iNatID = min(iNatID), .groups = "drop")

  attach_inat <- function(species_file, out_file) {
    species <- read_csv(species_file, show_col_types = FALSE) |>
      mutate(combo_key = str_to_lower(str_squish(Genus_Species_combo))) |>
      left_join(links_one, by = c("SiteCode_Plot", "combo_key")) |>
      select(-combo_key)
    write_csv(species, out_file)
    cat(
      "Wrote ", out_file, " with ", sum(!is.na(species$iNatID)),
      " iNatID values.\n", sep = ""
    )
    invisible(species)
  }

  attach_inat(plot_species_file, plot_inat_file)
  attach_inat(quad_species_file, quad_inat_file)
}
```
