---
title: "BioSCape Vegetation Plots, Cape Floristic Region, South Africa, 2023"
author: "Jasper Slingsby, Ross Turner, Douglas Euston-Brown, Greg Nicolson, Steven Molteno, Stuart Hall, Paul Emms,  Henry. A. Frye, Phillip. A. Townsend, Anabelle Cardoso, Erin Hestir, Brian Maitner, Adam M. Wilson"
date: today
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
  opts_knit:
    root.dir: ".."
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Paths, kept together so a new compilation tag is a one-line change.
library(tidyverse)
library(sf)
library(knitr)

source_tag <- "v20241104"
spatial_source_tag <- "v20261002"
release_tag <- "v20261002"
plot_summary_file <- "data/bioscape_veg_plot_summary_v20241104.csv"
plot_species_file <- "data/bioscape_veg_plot_species_v20241104.csv"
quadrat_summary_file <- "data/bioscape_veg_quadrat_summary_v20241104.csv"
quadrat_species_file <- "data/bioscape_veg_quadrat_species_v20241104.csv"
spatial_file <- "data/spatial/bioscape_veg_spatial_v20261002.gpkg"
out_dir <- "ornl_daac"

dataset_title <- paste(
  "BioSCape Vegetation Plots, Cape Floristic Region,",
  "South Africa, 2023"
)
```

```{r}
plot_summary <- read_csv(plot_summary_file, show_col_types = FALSE)
plot_species <- read_csv(plot_species_file, show_col_types = FALSE)
quadrat_summary <- read_csv(
  quadrat_summary_file,
  show_col_types = FALSE,
  col_types = cols(Date = col_datetime(format = ""))
)
quadrat_species <- read_csv(quadrat_species_file, show_col_types = FALSE)

# Plot centers, parking points, and plot polygons. Coordinates that
# are not in the CSV files are in PlotCenters.
centers <- read_sf(spatial_file, layer = "PlotCenters") |>
  st_transform(4326)
parking <- read_sf(spatial_file, layer = "PlotParking")
polygons <- read_sf(spatial_file, layer = "PlotPolygons")
```

```{r}
# Archive names, missing-value codes, and a single row id.
snake_feature <- function(x) {
  x |>
    str_replace_all("([a-z0-9])([A-Z])", "\\1_\\2") |>
    str_replace_all("([A-Z]+)([A-Z][a-z])", "\\1_\\2") |>
    str_replace_all("([A-Za-z])([0-9])", "\\1_\\2") |>
    tolower() |>
    str_replace("^bio_scape", "bioscape")
}

as_yes_no <- function(x) {
  case_when(
    x %in% c("1", "yes", "Yes") ~ "yes",
    x %in% c("0", "no", "No") ~ "no",
    TRUE ~ NA_character_
  )
}

parse_gps_time <- function(x) {
  parsed <- dmy_hms(str_replace(x, " at ", " "), quiet = TRUE)
  format(parsed, "%Y-%m-%dT%H:%M:%S", tz = "UTC")
}

finish_table <- function(df) {
  df |>
    rename_with(snake_feature) |>
    mutate(
      across(where(is.numeric), ~ replace_na(.x, -9999)),
      across(where(is.character), ~ replace_na(.x, "NA"))
    ) |>
    mutate(record_id = row_number(), .before = 1)
}
```

```{r}
# Counts of problems in the source tables, before they are rewritten.
site_case <- quadrat_species |>
  distinct(Plot, species_site = SiteCode) |>
  inner_join(
    quadrat_summary |> distinct(Plot, summary_site = SiteCode),
    by = "Plot"
  ) |>
  filter(species_site != summary_site)

n_urls_removed <- quadrat_species |>
  filter(!is.na(NewSpecies), str_detect(NewSpecies, "http")) |>
  nrow()

n_quadrat_species_raw <- nrow(quadrat_species)
```

```{r}
# Species tables inherit SiteCode from the plot summary so joins match.
site_lookup <- plot_summary |>
  distinct(Plot, site_code_standard = SiteCode)

plot_species <- plot_species |>
  select(-SiteCode, -SiteCode_Plot) |>
  left_join(site_lookup, by = "Plot") |>
  rename(SiteCode = site_code_standard) |>
  mutate(SiteCode_Plot = paste(SiteCode, Plot, sep = "_")) |>
  relocate(SiteCode, SiteCode_Plot, .after = Plot)

quadrat_species <- quadrat_species |>
  select(-SiteCode, -SiteCode_Plot, -SiteCode_Plot_Quadrant) |>
  left_join(site_lookup, by = "Plot") |>
  rename(SiteCode = site_code_standard) |>
  mutate(
    SiteCode_Plot = paste(SiteCode, Plot, sep = "_"),
    SiteCode_Plot_Quadrant = paste(
      SiteCode, Plot, Quadrant, sep = "_"
    )
  ) |>
  relocate(
    SiteCode, SiteCode_Plot, SiteCode_Plot_Quadrant, Quadrant,
    .after = Plot
  )

stopifnot(!any(is.na(plot_species$SiteCode)))
stopifnot(!any(is.na(quadrat_species$SiteCode)))

# Identical species rows are repeats. Differing measurements are kept.
quadrat_species <- distinct(quadrat_species)
n_exact_dups_removed <- n_quadrat_species_raw - nrow(quadrat_species)

plot_summary <- plot_summary |>
  mutate(
    Observer = if_else(
      Observer == "deb",
      "Douglas Euston-Brown",
      Observer
    )
  )

quadrat_summary <- quadrat_summary |>
  mutate(
    Date = as.Date(Date),
    Observer = if_else(
      Observer == "deb",
      "Douglas Euston-Brown",
      Observer
    ),
    sheet_name = str_extract(
      sheet_name,
      "Ross01c|Ross01b|Ross01|Bio02|Bio03|Bio04"
    ),
    PostFireAge_years = as.character(PostFireAge_years)
  )

plot_species <- plot_species |>
  mutate(Clonal = as_yes_no(as.character(Clonal)))

quadrat_species <- quadrat_species |>
  mutate(
    Clonal = as_yes_no(as.character(Clonal)),
    Clonal_YesNo = as_yes_no(as.character(Clonal_YesNo)),
    NameCheck = as.character(NameCheck),
    NameCheck = if_else(NameCheck %in% c("#N/A", "NA"), NA, NameCheck),
    NewSpecies = if_else(
      !is.na(NewSpecies) & str_detect(NewSpecies, "http"),
      NA_character_,
      as.character(NewSpecies)
    )
  )

plot_summary <- plot_summary |>
  arrange(Plot) |>
  finish_table()
plot_species <- plot_species |>
  arrange(Plot, desc(PercentCoverAlive)) |>
  finish_table()
quadrat_summary <- quadrat_summary |>
  arrange(Plot, Quadrant) |>
  finish_table()
quadrat_species <- quadrat_species |>
  arrange(Plot, Quadrant) |>
  finish_table()

# GPS clock times become ISO 8601. The source text has no timezone.
fix_layer <- function(x) {
  x |>
    mutate(
      across(any_of("DateTime"), parse_gps_time),
      across(
        any_of("Botanist"),
        ~ if_else(.x == "Doug Euston-Brown", "Douglas Euston-Brown", .x)
      )
    )
}

centers <- centers |>
  rename(any_of(c(PTPlotD = "PTPlotD.x"))) |>
  select(-any_of("PTPlotD.y")) |>
  fix_layer()
parking <- fix_layer(parking)
polygons <- fix_layer(polygons)

geom_name <- attr(centers, "sf_column")
rename_spatial <- function(x) {
  nm <- names(x)
  keep <- nm != attr(x, "sf_column")
  nm[keep] <- snake_feature(nm[keep])
  names(x) <- nm
  x
}
centers <- rename_spatial(centers)
parking <- rename_spatial(parking)
polygons <- rename_spatial(polygons)
```

```{r}
plot_summary_out <- file.path(
  out_dir, paste0("bioscape_veg_plot_summary_", release_tag, ".csv")
)
plot_species_out <- file.path(
  out_dir, paste0("bioscape_veg_plot_species_", release_tag, ".csv")
)
quadrat_summary_out <- file.path(
  out_dir, paste0("bioscape_veg_quadrat_summary_", release_tag, ".csv")
)
quadrat_species_out <- file.path(
  out_dir, paste0("bioscape_veg_quadrat_species_", release_tag, ".csv")
)
spatial_out <- file.path(
  out_dir, paste0("bioscape_veg_spatial_", release_tag, ".gpkg")
)

write_csv(plot_summary, plot_summary_out, na = "NA")
write_csv(plot_species, plot_species_out, na = "NA")
write_csv(quadrat_summary, quadrat_summary_out, na = "NA")
write_csv(quadrat_species, quadrat_species_out, na = "NA")

if (file.exists(spatial_out)) {
  file.remove(spatial_out)
}
st_write(centers, spatial_out, layer = "plot_centers", quiet = TRUE)
st_write(parking, spatial_out, layer = "plot_parking", quiet = TRUE)
st_write(polygons, spatial_out, layer = "plot_polygons", quiet = TRUE)
```

```{r}
# Every released plot was surveyed as four quadrants on one date.
quadrats_per_plot <- quadrat_summary |>
  count(site_code_plot, name = "n_quadrats")

dates_per_plot <- quadrat_summary |>
  distinct(plot, date) |>
  count(plot, name = "n_dates")

stopifnot(all(quadrats_per_plot$n_quadrats == 4))
stopifnot(all(dates_per_plot$n_dates == 1))

n_plots <- n_distinct(plot_summary$plot)
n_quadrats <- nrow(quadrat_summary)
n_plot_species <- nrow(plot_species)
n_quadrat_species <- nrow(quadrat_species)
n_taxa <- n_distinct(quadrat_species$genus_species_combo)

survey_start <- format(min(quadrat_summary$date), "%Y-%m-%d")
survey_end <- format(max(quadrat_summary$date), "%Y-%m-%d")

site_codes <- plot_summary |>
  distinct(site_code) |>
  arrange(tolower(site_code)) |>
  pull(site_code)

cover_off <- quadrat_summary |>
  filter(
    total_cover_check != 100,
    total_cover_check != -9999
  )

dup_taxon <- quadrat_species |>
  count(
    site_code_plot_quadrant,
    genus_species_combo,
    plot,
    site_code
  ) |>
  filter(n > 1)

namecheck_missing <- quadrat_species |>
  filter(name_check == "NA") |>
  nrow()

bbox <- st_bbox(centers)
centers_only <- centers |>
  st_drop_geometry() |>
  filter(!bioscape_plot_id %in% plot_summary$plot)

duplicate_centers <- centers |>
  st_drop_geometry() |>
  count(bioscape_plot_id, name = "n_points") |>
  filter(n_points > 1)

quality_flags <- centers |>
  st_drop_geometry() |>
  count(quality_flag, name = "Plots") |>
  filter(!is.na(quality_flag), quality_flag != "") |>
  arrange(desc(Plots))

center_repeat_note <- if (nrow(duplicate_centers) == 0) {
  "No plot id is repeated."
} else {
  paste0(
    paste(
      paste0(
        duplicate_centers$bioscape_plot_id,
        " is stored ",
        duplicate_centers$n_points,
        " times"
      ),
      collapse = "; "
    ),
    "."
  )
}

centers_missing_note <- if (nrow(centers_only) == 0) {
  "none"
} else {
  paste(centers_only$bioscape_plot_id, collapse = ", ")
}
```

```{r}
# Checksums identify the exact release files ORNL will archive.
release_files <- c(
  plot_summary_out,
  plot_species_out,
  quadrat_summary_out,
  quadrat_species_out,
  spatial_out
)
file_info <- tibble(
  file = release_files,
  rows = c(
    nrow(plot_summary),
    nrow(plot_species),
    nrow(quadrat_summary),
    nrow(quadrat_species),
    nrow(centers) + nrow(parking) + nrow(polygons)
  )
) |>
  mutate(
    bytes = file.size(file),
    sha256 = map_chr(file, function(path) {
      line <- system2("shasum", c("-a", "256", path), stdout = TRUE)
      str_split(line, "\\s+", n = 2)[[1]][1]
    }),
    file = basename(file)
  )
```


## Introduction

**Dataset title:** `r dataset_title`

**Summary:** Vegetation composition and cover at `r n_plots` field plots
in the Greater Cape Floristic Region of South Africa, collected for the
Biodiversity Survey of the Cape (BioSCape). Each plot was surveyed once
between `r survey_start` and `r survey_end` and divided into four
quadrants (NE, NW, SE, SW). The product reports quadrant-level and
plot-level fractional cover of bare soil, bare rock, dead vegetation,
and live vegetation, together with species-level percent cover, stem
counts, and canopy diameter for `r n_taxa` species. All species with at least 5% cover were included in the dataset, plus additional species where needed to make up 80% of the total cover. The data were collected in the field by botanists. 
Plot-center coordinates are in the companion spatial file. 

## Dataset overview

Project: Biodiversity Survey of the Cape (BioSCape), an integrated field
and airborne campaign in the Greater Cape Floristic Region (Cardoso et
al. 2025).

**Related publication:**

Cardoso, A.W., E.L. Hestir, J.A. Slingsby, C.J. Forbes, G.R. Moncrieff,
W. Turner, A.L. Skowno, J. Nesslage, P.G. Brodrick, K.D. Gaddis, and
A.M. Wilson. 2025. The biodiversity survey of the Cape (BioSCape),
integrating remote sensing with biodiversity science. *npj
Biodiversity* 4:2. <https://doi.org/10.1038/s44185-024-00071-5>


**Related datasets:**

- Fitzpatrick, M.C., D. Spalink, A.J. Elmore, J. Measey, and
  S. Kritzinger-Klopper. 2026. BioSCape: Vegetation Composition at Plots
  with Woody Invasives, CFR, South Africa. ORNL DAAC.
  <https://doi.org/10.3334/ORNLDAAC/2511>. Those plots follow this
  survey protocol and were placed near these plots, in vegetation that
  also contained woody invasive cover.
- Slimp, M., J. Malindi, R. Meyer, A. Cunningham, J. Rourke, M. Hayden,
  A. Hargey, J. Nesslage, S. Johnson, M. Rossi, and N. Stavros. 2025.
  BioSCape Vegetation Surveys Berg and Eerste River Catchments, South
  Africa, 2022–2023. ORNL DAAC.
  <https://doi.org/10.3334/ORNLDAAC/2425>. Same campaign, different
  sites and some different measurements.

**Acknowledgments:** Cardoso et al. (2025) report NASA award
80NSSC21K0086 to A.M. Wilson and E.L. Hestir for the U.S. components of
BioSCape, and National Research Foundation grant 142438 to J.A. Slingsby. 

## Dataset characteristics

**Spatial coverage:** Greater Cape Floristic Region, Western Cape and
Eastern Cape, South Africa. Site codes in the plot summary:
`r paste(site_codes, collapse = ", ")`.

**Spatial resolution:** One circular vegetation plot with a diameter of 10m at each location, subdivided
into four quadrants. 

**Temporal coverage:** `r survey_start` 00:00:00 to `r survey_end`
23:59:59 UTC. `date` in the quadrat summary is the survey calendar date.
Each plot has one date.

**Temporal resolution:** One sample per plot.

**Study area:** Bounding box of the `plot_centers` layer in
`r basename(spatial_out)`, WGS 84 decimal degrees.

```{r}
tibble(
  Site = "Greater Cape Floristic Region, South Africa",
  `Westernmost longitude` = sprintf("%.4f", bbox[["xmin"]]),
  `Easternmost longitude` = sprintf("%.4f", bbox[["xmax"]]),
  `Northernmost latitude` = sprintf("%.4f", bbox[["ymax"]]),
  `Southernmost latitude` = sprintf("%.4f", bbox[["ymin"]])
) |>
  kable(format = "pipe")
```

**GCMD keywords:** BIOSPHERE > VEGETATION > VEGETATION COVER; BIOSPHERE >
VEGETATION > VEGETATION SPECIES; AFRICA > SOUTHERN AFRICA > SOUTH
AFRICA; platform FIELD SURVEYS; instrument GPS.

### File naming convention

`bioscape_veg_{level}_{table}_v{YYYYMMDD}.csv` and
`bioscape_veg_spatial_v{YYYYMMDD}.gpkg`

- `level` is `plot` or `quadrat`.
- `table` is `summary` (one row per location) or `species` (one row per
  taxon at that location).
- `YYYYMMDD` is the date these version 1 files were written. This
  submission is `r release_tag`. The vegetation sources are
  `r source_tag` and previously released to the team on GitHub. The spatial source is `r spatial_source_tag`.
  Publication at ORNL is version 1. There is no earlier DOI.

Names use lowercase letters, numbers, and underscores.

### File descriptions

The CSV files have one header row. Numeric missing values are `-9999`.
Text missing values are `NA`. `record_id` is unique within each CSV.

```{r}
file_info |>
  kable(format = "pipe")
```

- **Plot summary** (`r n_plots` rows). One row per surveyed plot.
  Fractional cover, vegetation height, and soil depth are means of the
  four quadrants. `site_code_plot` is unique.
- **Plot species** (`r n_plot_species` rows). One row per accepted taxon
  per plot. Live and dead cover are the sum of the quadrant percent
  covers divided by 4. Counts are sums across quadrants.
  `site_code_plot` plus `genus_species_combo` is unique except where
  noted in Quality assessment.
- **Quadrat summary** (`r n_quadrats` rows). One row per quadrant.
  `site_code_plot_quadrant` is unique. All `r n_plots` plots have NE,
  NW, SE, and SW.
- **Quadrat species** (`r n_quadrat_species` rows). One row per taxon
  recorded in a quadrant, after identical rows were dropped. The plot
  species file is the aggregate of the source table, not of this file
  after duplicate removal.

**Spatial file.** `r basename(spatial_out)` is a GeoPackage in WGS 84
(EPSG:4326). Layers are `plot_centers`, `plot_parking`, and
`plot_polygons`.

- `plot_centers`, `r nrow(centers)` points and
  `r n_distinct(centers$bioscape_plot_id)` plot ids.
  `r center_repeat_note` Center ids absent from the plot summary:
  `r centers_missing_note`. `bioscape_plot_id` matches `plot`.
- `plot_parking`, `r nrow(parking)` points where a vehicle was parked
  for the survey.
- `plot_polygons`, `r nrow(polygons)` Patches of vegetation the botanists deemed to be representative of the plot (e.g. similar species, post-fire age, soil) that are at least 15 meters in radius and ideally over 50 meters.

`plot_centers` attributes:

| Variable | Description |
| --- | --- |
| bioscape_plot_id | Name of the BioSCape plot. Matches `plot`. |
| region | General region where the plot is located. |
| botanist | Name of the botanist who conducted the vegetation survey. |
| date_time | Clock time of the GPS measurement, `YYYY-MM-DDThh:mm:ss`. The source text did not include a timezone. |
| name | Name the botanist gave the plot, kept for provenance. |
| plot_note | Notes made by the botanist, kept for provenance. |
| description | Short free-text note. |
| quality_flag | Flag assigned by the Townsend team while visiting the plot during BioSCape. Values are listed below. |
| quality_flag_2 | A second Townsend visit flag. One plot is `Boundary`. |
| location_flag | The original GPS may not have been taken at the plot center. This flag is set when that plot's BioSCape center and its own Townsend center differ by more than 10 m. The stored value is `See Townsend alternative location`. See `townsend_long` and `townsend_lat`. |
| pt_plot_a | Townsend opportunistic plot associated with this BioSCape plot. Additional plots were collected where the original plot had a quality flag, where nearby vegetation was similar but had a different dominant species, or where a nearby patch was a single species. |
| pt_plot_b | A second Townsend plot label. See `pt_plot_a`. |
| pt_plot_c | A third Townsend plot label. See `pt_plot_a`. |
| pt_plot_d | A fourth Townsend plot label. See `pt_plot_a`. |
| townsend_long | Alternative longitude, decimal degrees, of the plot center marked by the Townsend team. |
| townsend_lat | Alternative latitude, decimal degrees, of the plot center marked by the Townsend team. |
| geom | Longitude and latitude of the BioSCape plot center. |

`quality_flag` values in `plot_centers`:

```{r}
quality_flags |>
  kable(format = "pipe")
```

`Boundary` means the plot was placed on the boundary of two vegetation types. The Townsend team then set up additional plots nearby to record the distinct communities (`pt_plot_a` through `pt_plot_d`). `Low vegetation` means cover was very low, and an alternative plot was often marked nearby where cover was higher. `Trampled` means previous visitors had trampled the plot. `Removed invasive` means a Townsend-team botanist removed an invasive individual that could have been present during the flights. `Grazing` means there was evidence of grazing. An invaded flag means invasive or alien plants were present, and a recent-burn flag means there was evidence of recent fire. Neither of those two labels is used in this file. `Other` and `Accessibility` are stored on some plots.

`plot_parking` columns are `bioscape_plot_id`, `region`, `name`, `description`, and `date_time`. `plot_polygons` columns are `bioscape_plot_id`, `region`, `botanist`, `name`, `description`, and `date_time`.

**Code.** `workflow/plot_data_compile_xlsx.R` in
<https://github.com/BioSCape-io/BioSCape-terrestrial>. Released tables:
<https://github.com/BioSCape-io/BioSCape-terrestrial/releases/tag/v20241104>.

### Data file properties

Tabular CSV, plus one GeoPackage. The four tables have no map
projection. `r basename(spatial_out)` uses geographic WGS 84
(EPSG:4326). Processing level: in situ field
observations, with accepted names applied at compilation. These are not
a satellite Level 2–4 product.

### Data details

Shared codes:

- `quadrant`: NE, NW, SE, SW.
- `clonal` and `clonal_yes_no`: `yes`, `no`, or `NA`.
- `seasonally_apparent`: `0`, `1`, or `-9999`.
- `groundwater`: Well-drained, Impeded drainage, Seepage, Swamp, Stream
  bank, Suurvlakte, or `NA`.
- `sampled` on the plot summary is `1` for every row.

**`r basename(plot_summary_out)`**

| Variable | Units | Description |
| --- | --- | --- |
| record_id | | Unique row number in this file. |
| plot | | Plot code, `T` plus three digits. Unique in this file. |
| site_code | | Survey area name. |
| site_code_plot | | Site and plot joined by `_`. Unique row key. |
| gps_plot_centre | | Text note on how the center was recorded. Not a coordinate. |
| observer | | Botanist named on the field sheet. `deb` is stored as Douglas Euston-Brown. |
| percent_bare_soil | percent | Mean of the four quadrant estimates. Missing: `-9999`. |
| percent_bare_rock | percent | Mean of the four quadrant estimates. Missing: `-9999`. |
| percent_dead_vegetation | percent | Mean of the four quadrant estimates. Missing: `-9999`. |
| percent_live_vegetation | percent | Mean of the four quadrant estimates. Missing: `-9999`. |
| veg_height_mean_cm | cm | Mean vegetation height across quadrants. Missing: `-9999`. |
| soil_depth_cm | cm | Mean soil depth across quadrants. Missing: `-9999`. |
| groundwater | | Drainage class taken from the quadrants. |
| access_notes | | Free-text access note. |
| sampled | | Completed-survey flag. Always 1. |

This file has no survey date. Take `date` from the quadrat summary,
joining on `plot`. Field surveys were recorded by Ross Turner, Douglas
Euston-Brown, Steven Molteno, Stuart Hall, and Paul Emms.

**`r basename(plot_species_out)`**

| Variable | Units | Description |
| --- | --- | --- |
| record_id | | Unique row number in this file. |
| plot | | Plot code. |
| site_code | | Survey area name. |
| site_code_plot | | Site and plot joined by `_`. |
| accepted_genus | | Accepted genus. |
| accepted_species | | Accepted species epithet. |
| name_check | | Name-check code copied from the field sheet. The list that issued the numbers is not identified in the source workbooks. Missing: `NA`. |
| genus_species_combo | | Accepted genus and species as one label. |
| percent_cover_alive | percent | Sum of quadrant live cover, divided by 4. Missing: `-9999`. |
| percent_cover_dead | percent | Sum of quadrant dead cover, divided by 4. Missing: `-9999`. |
| abundance_alive_count | count | Sum of live counts across quadrants. Missing: `-9999`. |
| abundance_dead_count | count | Sum of dead counts across quadrants. Missing: `-9999`. |
| mean_canopy_diameter_cm | cm | Mean of recorded quadrant canopy diameters. Missing: `-9999`. |
| clonal | | `yes`, `no`, or `NA`. |
| seasonally_apparent | | Maximum of the quadrant values. Missing: `-9999`. |

**`r basename(quadrat_summary_out)`**

| Variable | Units | Description |
| --- | --- | --- |
| record_id | | Unique row number in this file. |
| plot | | Plot code. |
| site_code | | Survey area name. |
| site_code_plot | | Site and plot joined by `_`. |
| site_code_plot_quadrant | | Site, plot, and quadrant joined by `_`. Unique row key. |
| quadrant | | NE, NW, SE, or SW. |
| gps_plot_centre | | Text note. Not a coordinate. |
| observer | | Botanist named on the field sheet. |
| date | | Survey date, `YYYY-MM-DD`. |
| percent_bare_soil | percent | Quadrant estimate. Missing: `-9999`. |
| percent_bare_rock | percent | Quadrant estimate. Missing: `-9999`. |
| percent_dead_vegetation | percent | Quadrant estimate. Missing: `-9999`. |
| percent_live_vegetation | percent | Quadrant estimate. Missing: `-9999`. |
| total_cover_check | percent | Sum of the four cover fractions. `r nrow(cover_off)` quadrants are not 100 and are not missing. Missing: `-9999`. |
| veg_height_cm | cm | Vegetation height. Missing: `-9999`. |
| post_fire_age_years | text | Time since fire as written on the sheet. The column is text because some entries are notes or Excel date serials rather than ages. |
| soil_depth_cm | cm | Soil depth. Missing: `-9999`. |
| groundwater | | Drainage class. |
| parking_gps | | Text note that a parking location was recorded. Not a coordinate. |
| access_notes | | Free-text access note. |
| comments | | Free-text site description. |
| sheet_name | | Source workbook: Ross01, Ross01b, Ross01c, Bio02, Bio03, or Bio04. |
| old_plot | | Plot number before the Swartberg renumbering in Methods. Missing: `-9999`. |

**`r basename(quadrat_species_out)`**

| Variable | Units | Description |
| --- | --- | --- |
| record_id | | Unique row number in this file. |
| plot | | Plot code. |
| site_code | | Survey area name. |
| site_code_plot | | Site and plot joined by `_`. |
| site_code_plot_quadrant | | Site, plot, and quadrant joined by `_`. |
| quadrant | | NE, NW, SE, or SW. |
| accepted_genus | | Accepted genus. |
| accepted_species | | Accepted species epithet. `NA` when the record is genus only. |
| subspecies_or_variant | | Subspecies or variant when one was recorded. |
| genus_species_combo | | Genus and species label. |
| name_check | | Name-check code. `NA` on `r namecheck_missing` of `r n_quadrat_species` rows. |
| new_species | | Field note for a new or uncertain taxon. URLs were removed. Remaining values can be notes or times. |
| clonal_yes_no | | `yes`, `no`, or `NA`. |
| mean_canopy_diameter_cm | cm | Canopy diameter of the taxon in the quadrant. Missing: `-9999`. |
| abundance_alive_count | count | Live individuals. Missing: `-9999`. |
| percent_cover_alive | percent | Live cover in the quadrant. Missing: `-9999`. |
| percent_cover_dead | percent | Dead cover in the quadrant. Missing: `-9999`. |
| abundance_dead_count | count | Dead individuals. Missing: `-9999`. |
| seasonally_apparent | | `0`, `1`, or `-9999`. |
| comment | | Note on the taxon in that quadrant. |
| clonal | | `yes`, `no`, or `NA`. |
| taxon | | Same label as `genus_species_combo`. |
| old_plot | | Plot number before Swartberg renumbering. Missing: `-9999`. |


## Application and derivation

These plots are the environmentally stratified vegetation surveys for
BioSCape. They are the ground observations for comparing field
composition and structure with the campaign airborne imaging
spectroscopy and lidar (AVIRIS-NG, PRISM, LVIS, and HyTES) over the
Greater Cape Floristic Region (Cardoso et al. 2025). Use the plot
species table for one cover and count per taxon per plot. Use the
quadrat tables when the within-plot measurements matter. The
woody-invasive plots (ORNL DAAC
<https://doi.org/10.3334/ORNLDAAC/2511>) were sited near this network
and are comparable only after that placement difference is taken into
account.

## Quality assessment

Cover fractions are visual estimates. This version does not add an
uncertainty column. `total_cover_check` is the field sum of bare soil,
bare rock, dead vegetation, and live vegetation. Quadrants that do not
sum to 100:

```{r}
cover_off |>
  select(site_code_plot_quadrant, total_cover_check) |>
  arrange(site_code_plot_quadrant) |>
  kable(format = "pipe")
```

Version 1 changes from the source tables:

- Site codes on `r nrow(site_case)` plots in the species tables were
  replaced with the plot-summary spelling so `site_code_plot` joins.
- `r n_exact_dups_removed` identical quadrat-species rows were removed.
- `r n_urls_removed` web address in `new_species` was removed.
- The observer value `deb` and the spatial botanist name
  `Doug Euston-Brown` are stored as Douglas Euston-Brown.
- Spatial source `r spatial_source_tag` drops the duplicate T114
  center and its location flag, sets T074 region to `CapePoint`,
  links T183 to associated plot `pt183`, stores `pt_plot_d` as one
  column, and uses real missing values instead of the text `NA` in
  Townsend plot labels.

`r nrow(dup_taxon)` quadrat-by-taxon keys still occur more than once
because the cover or count differed. `record_id` keeps those rows
distinct. The plot species file was summed from the source table, so
those repeated measurements are included in its cover and counts.

```{r}
dup_taxon |>
  arrange(site_code_plot_quadrant, genus_species_combo) |>
  select(site_code_plot_quadrant, genus_species_combo, n) |>
  kable(format = "pipe")
```

`post_fire_age_years` remains text. `name_check` is the code written on
the field sheet; the authority list is not in these tables.

## Data acquisition, materials, and methods

Plots were placed across reserves and passes in the Greater Cape
Floristic Region for the 2023 BioSCape field campaign (Cardoso et al.
2025). Each surveyed plot has four quadrants: NE, NW, SE, and SW. In
each quadrant the botanist recorded fractional cover of bare soil, bare
rock, dead vegetation, and live vegetation; vegetation height; soil
depth; drainage; and, for each vascular plant taxon, live and dead
percent cover, live and dead counts, and canopy diameter. Centers were
located with a Trimble GPS. The coordinate values are in the
`plot_centers` layer of `r basename(spatial_out)`.

The version 1 files were written by this document. Column names are
snake case. Numeric gaps are `-9999` and text gaps are `NA`. Survey
dates are `YYYY-MM-DD`. GPS times are `YYYY-MM-DDThh:mm:ss` with the
clock time taken from the source string. Each CSV has `record_id`.
Clonal codes are `yes` or `no`. `old_plot` is the plot number before
renumbering.

Field records were kept in six Google Sheets workbooks (Ross01,
Ross01b, Ross01c, Bio02, Bio03, Bio04) and downloaded on 2024-11-04.
`workflow/plot_data_compile_xlsx.R` stacked the SiteData tabs and the
plot tabs, and rebuilt the plot and quadrant identifiers. Swartberg plot numbers
that collided with numbers used at other sites were renumbered. The
previous number is `old_plot`. The source script maps Swartberg 20 to T110, 22
to T012, 23 to T013, and 24 to T014.

Plot summary cover, height, and soil depth are the mean of the four
quadrants. Plot species live and dead cover are the sum of that taxon's
quadrant covers divided by 4, and counts are sums. Unrecorded quadrant
occurrences add nothing to the sum. With four quadrants in every
released plot, dividing by 4 is that taxon's mean contribution to the
plot.

## References

Cardoso, A.W., E.L. Hestir, J.A. Slingsby, C.J. Forbes, G.R. Moncrieff,
W. Turner, A.L. Skowno, J. Nesslage, P.G. Brodrick, K.D. Gaddis, and
A.M. Wilson. 2025. The biodiversity survey of the Cape (BioSCape),
integrating remote sensing with biodiversity science. *npj
Biodiversity* 4:2. <https://doi.org/10.1038/s44185-024-00071-5>

Fitzpatrick, M.C., D. Spalink, A.J. Elmore, J. Measey, and
S. Kritzinger-Klopper. 2026. BioSCape: Vegetation Composition at Plots
with Woody Invasives, CFR, South Africa. ORNL DAAC, Oak Ridge,
Tennessee, USA. <https://doi.org/10.3334/ORNLDAAC/2511>

Slimp, M., J. Malindi, R. Meyer, A. Cunningham, J. Rourke, M. Hayden,
A. Hargey, J. Nesslage, S. Johnson, M. Rossi, and N. Stavros. 2025.
BioSCape Vegetation Surveys Berg and Eerste River Catchments, South
Africa, 2022–2023. ORNL DAAC, Oak Ridge, Tennessee, USA.
<https://doi.org/10.3334/ORNLDAAC/2425>

## Dataset revisions

 This release is compiled from `r source_tag` vegetation tables and the `r spatial_source_tag` spatial
GeoPackage from the team GitHub repository. It was updated to conform to ORNL data standards as `r release_tag`. 
This is the first submission of these tables to ORNL DAAC. There is no prior DOI.

```{r}
#| eval: false
#| include: false

# Publish the version 1 files to GitHub. Render this document first so
# README.md exists, then run this chunk with the working directory set
# to the repository root. A normal render does not upload.
library(piggyback)

github_repo <- "BioSCape-io/BioSCape-terrestrial"
github_tag <- "ORNLv1"
release_tag <- "v20261002"
github_dir <- "ornl_daac"
github_files <- c(
  file.path(
    github_dir,
    paste0("bioscape_veg_plot_summary_", release_tag, ".csv")
  ),
  file.path(
    github_dir,
    paste0("bioscape_veg_plot_species_", release_tag, ".csv")
  ),
  file.path(
    github_dir,
    paste0("bioscape_veg_quadrat_summary_", release_tag, ".csv")
  ),
  file.path(
    github_dir,
    paste0("bioscape_veg_quadrat_species_", release_tag, ".csv")
  ),
  file.path(
    github_dir,
    paste0("bioscape_veg_spatial_", release_tag, ".gpkg")
  ),
  file.path(github_dir, "README.md")
)
stopifnot(all(file.exists(github_files)))

# pb_releases() caches its result for the R session. A release created
# earlier in the session stays invisible until that cache is cleared.
memoise::forget(piggyback::pb_releases)
memoise::forget(piggyback::pb_info)

if (!github_tag %in% pb_releases(repo = github_repo)$tag_name) {
  pb_new_release(repo = github_repo, tag = github_tag)
  memoise::forget(piggyback::pb_releases)
  memoise::forget(piggyback::pb_info)
}

pb_upload(
  file = github_files[1],
  repo = github_repo,
  tag = github_tag,
  overwrite = TRUE
)
pb_upload(
  file = github_files[2],
  repo = github_repo,
  tag = github_tag,
  overwrite = TRUE
)
pb_upload(
  file = github_files[3],
  repo = github_repo,
  tag = github_tag,
  overwrite = TRUE
)
pb_upload(
  file = github_files[4],
  repo = github_repo,
  tag = github_tag,
  overwrite = TRUE
)
pb_upload(
  file = github_files[5],
  repo = github_repo,
  tag = github_tag,
  overwrite = TRUE
)
pb_upload(
  file = github_files[6],
  repo = github_repo,
  tag = github_tag,
  overwrite = TRUE
)
```
